How to select predictive models for decision-making or causal inference?

### 2024-10-19 response_to_reviewer.md

We thank the reviewers for their time, thoughts, and feedback on the
manuscript. We have addresed the comments through additional work as well
as clarifications in the revised manuscript.

@@

# Reviewer 1

*[...] The authors conceivably state that "The problem is that causal estimation requires controlling an error on both treated and non-treated outcome for the same individual.". Consequently, they report on estimator that attempt to consider this (in Section "Methods"). However, I cannot understand how this hint ("controlling error on both treated and non-treated sample (or patient?)") has been implemented by the procedure the authors describe in Equations (1), (2), and (3). If, I correctly understand", in Eq. (2) we see the g-computation estimators, a well-known concept introduced by [21] (and recently used also by https://www.nature.com/articles/s41598-020-65917-x ), that splits the two configurations "treated" and "non-treated". [...] I understand that Eq. (2) and in Eq. (3) consider both treated and non-treated sample (or patient?), but those sums can be also equal to 0, even if their single terms are different from 0. Moreover, it is not clear if, and how Eq. (3) is related to Eq. (2): is it a variant? Is Eq. (3) derived from Eq. (2) under special assumptions?*

> The reviewer is right that Eq. (2) is part of constructing the classic
> g-computation estimator, as well as Eq. (1) and Eq. (3). We added a
> sentence at the begining of the section to stress that it is a section
> summarizing a classic framework.

> Eq. (3) is the definition of the estimator of the ATE based on a model of the outcome. It is sometimes called g-estimation. It  is unbiased under under consistency (Assumption. 3) and unconfoundedness (Assumption 1). To prove that, take the expectation of Eq. (3).  Then apply  Eq. (1) which is true under consistency (Assumption 3) and unconfoundedness (Assumption 1). Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq. (2).

> G-estimation itself does not control error on both populations as highlighted in Figure 1. Specific inductive biases of ML models are able to favor such regularization(see for example [Johannson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf)). Selection procedures, more elaborated than RMSE are needed to favor such models. I added this last explanation in the method (par. Model-selection risks, oracle and feasible) trying to clarify the link between the hint on error control on both population, g-estimation and the selection procedure which is our main focus.

### 2024-10-19 response_to_reviewer.md

# Reviewer 1

*[...] The authors conceivably state that "The problem is that causal estimation requires controlling an error on both treated and non-treated outcome for the same individual.". Consequently, they report on estimator that attempt to consider this (in Section "Methods"). However, I cannot understand how this hint ("controlling error on both treated and non-treated sample (or patient?)") has been implemented by the procedure the authors describe in Equations (1), (2), and (3). If, I correctly understand", in Eq. (2) we see the g-computation estimators, a well-known concept introduced by [21] (and recently used also by https://www.nature.com/articles/s41598-020-65917-x ), that splits the two configurations "treated" and "non-treated". [...] I understand that Eq. (2) and in Eq. (3) consider both treated and non-treated sample (or patient?), but those sums can be also equal to 0, even if their single terms are different from 0. Moreover, it is not clear if, and how Eq. (3) is related to Eq. (2): is it a variant? Is Eq. (3) derived from Eq. (2) under special assumptions?*

> The reviewer is right that Eq. (2) is part of constructing the classic
> g-computation estimator, as well as Eq. (1) and Eq. (3). We added a
> sentence at the begining of the section to stress that it is a section
> summarizing a classic framework.

> Eq. (3) is an expression of the estimator of the ATE based on a model of the outcome, sometimes called g-estimation. It is unbiased under consistency (Assumption. 3) and unconfoundedness (Assumption 1). To prove that, take the expectation of Eq. (3).  Then apply  Eq. (1) which is true under consistency (Assumption 3) and unconfoundedness (Assumption 1). Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq. (2).

> G-estimation itself does not control error on both populations. The formulas in Eq. (1-3) are all partly oracle formulas: they rely on conditional expectations but give no specific procedures on how to compute them. We added a clarification on this topic above Eq. (4) and above Eq. (8).

### 2024-10-19 how_to_select_latex.tex
iter

@@

Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
finite sum version of \autoref{eq:tau_population} leads to an unbiased
estimator of the ATE written:
\begin{equation}
    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
    \label{eq:ate_estimate}
\end{equation}
This estimator is an oracle \emph{finite sum} estimator by opposition to the
population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
$,
which involves an expectation taken on the full
distribution $\mathcal D$, which is observable but requires infinite data. For
each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
$\hat \ell$ to note its finite sum version. In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.

@@

\paragraph{Robinson decomposition}
G-computation is a choice of decomposition of the CATE estimation into a
sequence of regressions, where the only statistical estimate is the response
function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices
of decomposition exist, such as the R-decomposition
\cite{robinson_rootnconsistent_1988}. The latter introduces two new statistical
estimates, the conditional mean outcome and the probability of being treated (known
as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).

@@

We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
However, g-computation itself does not control error on both populations as highlighted in Figure \ref{fig:toy_example}. Specific inductive biases of ML models are able to favor such regularization \cite{johansson2022generalization}. Thus, selection procedures, more elaborated than MSE are needed to favor the best model for each dataset. A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ with observed data $O$ and a splitting strategy of the data for estimating the different regressions (nuisances) involved in the risk.
%
Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}

>>>>

We first expose the classic construction of the outcome modeling (or
G-computation) estimators of causal effect
\cite{robins_new_1986,snowden_implementation_2011,hernan_causal_2020}.

@@

Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
finite sum version of \autoref{eq:tau_population} leads to an unbiased
estimator of the ATE written:
\begin{equation}
    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
    \label{eq:ate_estimate}
\end{equation}
This estimator is an oracle \emph{finite sum} estimator by opposition to the
population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
$,
which involves an expectation taken on the full
distribution $\mathcal D$, which is observable but requires infinite data. For
each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
$\hat \ell$ to note its finite sum version. 
The formulas in Eq. (1-3) are all partly oracle formulas: they rely on
conditional expectations, the response functions, but give no specific
procedures on how to compute or select them. This last point is the topic
of our work, describe in the next section.
%In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.

@@

\paragraph{Robinson decomposition}
G-computation is a choice of decomposition of the CATE estimation. Other choices
of decomposition exist, such as the R-decomposition
\cite{robinson_rootnconsistent_1988}. The latter introduces two new statistical
estimates, the conditional mean outcome and the probability of being treated (known
as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).

@@

We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
However, the g-computation framework presented above is written in terms
of ``perfect'' conditional expectations (oracles), it does not control an
error, \emph{eg} on both populations as highlighted in Figure
\ref{fig:toy_example}. Selection procedures are needed to favor the best model for each dataset.

A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ with observed data $O$ and a splitting strategy of the data for estimating the different regressions (nuisances) involved in the risk.
%
Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}
